Dynamic capabilities and competitive advantage

### 2019-10-31 article4_jbr.rnw
almost finished first draft of hyp lacking only hyp 5

Looking at dynamic capabilities through the functions of sensing, seizing and
transformation \cite{Teece2007} makes an obvious case for resource orchestration which is also highlighted
in the extant literature. Moreover, as discussed above, these functions are also requiring
a certain non-behavioral object of action in order to be fully efficient. To achieve a
competitive edge firms will have to be able to sense and seize, which require a certain
organizational cognitive frame to be efficient. 'Dynamic capabilities are about doing the
right things, at the right time, based on new product (and process) development, unique
managerial orchestration processes, a strong and {\bf change-oriented organizational culture}'
(our emphasis) \cite[p]{Teece2014}.

To see how dynamic capabilities acts through cognitive objects of action (i.e. as a
cognitive programmer) we consider two distinctly different functions of dynamic
capabilities related to their microfoundations \citep{Teece2007}. the first function is
related to the discovering and capturing opportunities (e.g. the ability of the
organization to sense and seize opportunities (ibid)) which is very much related to the
concept of exploration \citep{March1991}. Organizations with high levels of dynamic
capabilities geared towards sensing and seizing opportunities are doing so through
\emph{resource orchestration}, e.g. by creating, extending and modifying the resource
base, as discussed extensively in the extant literature
\citep{Helfat2007,DiStefano2014,Protogerou2012,Schilke2018}. However, sensing and seizing
opportunities in the market relates to discovering market disequilibrium
\citep{Kirzner1997} as well as upsetting existing equilibria \citep{Schumpeter1934}. The
process of sensing and seizing 'involves understanding latent demand, the structural evolution of
industries and markets, and likely supplier and competitor responses'
\citep[p. 1322]{Teece2007} as well as . Individuals are the unit of analysis embedding these abilities
and if left up to these individuals without a higher organizing framework, the firm will
be very vulnerable because of knowledge asymmetries and incentive problems (alchian and
demsetz 1972) and lead to sub optimal rent-appropriation on the part of the individual
employee \citep{Blyler2003}. A similar argument applies to the function of \emph{seizing}
where the organization is set to mobilize resources for the purpose of gaining from these
opportunities. In order for value to be captured from opportunities, a string of
conditions have to be met \citep{Teece2017}, but the evolution of a 'strong and
change-oriented organizational culture' \citep[p. 331]{Teece2014} is of pivotal
importance. Alas the sensing process requires new knowledge from a wide range of sources
in the organization a lot of this knowledge is tacit and can lead to rent-appropriation
from information asymmetries \citep{Blyler2003} if not sufficiently aligned with the
cognitive frames of the organization.

@@

@@

The concept of sensing and seizing has analogous meaning in theoretical traditions tangent
to DC. Notably entrepreneurship \citep{Hitt2011} and organizational theory
\citep{March1991}. \cite{Hitt2011} states that "entrepreneurship involves identifying and
exploiting opportunities" (ibid p 59) where identification is tied to the process of
discovery \citep{Shane2000}. Based on the definitions of opportunity discovery as DEF HERE
juxtaposed to the definition of sensing as "a scanning, creation, learning, and
interpretive activity" \citet[p. 1322]{Teece2007} we argue that they are similar in nature
and usable for exploring the importance of the latter in the innovation process. It also
bears resemblance to the concept of organizational exploration as searching, risk-taking,
discovery, experimentation, prototyping, and flexibility \citep{March1991}. Independent of
the nature of opportunities to be sensed as being either objective
\citep{Alvarez2007,Shane2000} or created \citep{Alvarez2007}, the process of sensing does
require resources and capabilities for search (ref) and a clear affinity between
opportunity and a project exist (casson and Wadeson in Mcmullen2007) suggesting s clear
inter dependency between routines and resources, and higher order capabilities of sensing
\citep{Teece2007} and making sense of the opportunity
\citep{Weick1995,March1991}. Consequently, we would expect the sensing class of dynamic
capabilities to be positively related to innovation on the one hand, and reinforcing the
impact from innovation input simultaneously.\newline

>>>>

Looking at dynamic capabilities through the functions of sensing and seizing opportunities
\cite{Teece2007} makes an obvious case for resource orchestration which is also
highlighted in the extant literature. This function is related to the discovering and
capturing opportunities (e.g. the ability of the organization to sense and seize
opportunities (ibid)) which is very much related to the concept of exploration
\citep{March1991}Moreover, as discussed above, these functions are also requiring a
certain non-behavioral object of action in order to be fully efficient. To achieve a
competitive edge firms will have to be able to sense and seize, which require a certain
organizational cognitive frame to be efficient. 'Dynamic capabilities are about doing the
right things, at the right time, based on new product (and process) development, unique
managerial orchestration processes, a strong and {\bf change-oriented organizational
  culture}' (our emphasis) \cite[p]{Teece2014}.

changes resulting from execution of dynamic capabilities (e.g. sensing and seizing) may
move existing organizational objective functions and thus render current incentive
alignment obsolete or out of place. This leads to an objective displacement that is
particularly prevalent in firms with high sensing and seizing capabilities because distant
search may lead to more radical change (ref). Organizations with high levels of dynamic
capabilities geared towards sensing and seizing opportunities are doing so through
\emph{resource orchestration}, e.g. by creating, extending and modifying the resource
base, as discussed extensively in the extant literature
\citep{Helfat2007,DiStefano2014,Protogerou2012,Schilke2018}. However, sensing and seizing
opportunities in the market relates to discovering market disequilibrium
\citep{Kirzner1997} as well as upsetting existing equilibria \citep{Schumpeter1934}. The
process of sensing and seizing 'involves understanding latent demand, the structural
evolution of industries and markets, and likely supplier and competitor responses'
\citep[p. 1322]{Teece2007} as well as 'a good deal of customer, competitor, and supplier
information and intelligence (with a) a significant tacit component' (ibid p 1330).To make
sure the organization respond to changes (i.e. having an efficient cognitive frame)
stemming from executed dynamic capabilities (e.g. sensing and seizing) the firm needs to
make sure incentives are aligned for change.

Through the process of exploration, individuals in the organization may
accumulate private and tacit information that, if unchecked, can become a source of
knowledge asymmetries and incentive problems (alchian and demsetz 1972), and leave the
firm vulnerable to agency problems. Accumulation of private information in the
organization without incentive alignment of some sort can lead rent-appropriation from
information asymmetries \citep{Blyler2003}. In such instances of potential misalignment of
interest, a shared cognitive frame emphasizing a growth mindset can be a useful
remedy. Furthermore, in addition to incentives alignment the more psychological element
of change buy-in comes into play \todo{Not sure how to proceed here - should we make a
  distinction between the incentive problem and the buy in problem? Coud they both be
  examples of interest alignment with different faces of motivation - extrensic vs intrinsic} (ref).

To see how this would work recent contribution in neuroscience is helpful. Several
contributions have established a clear connection between having a growth mindset and
emphasizing intrinsic motivation (Schroder et.al 2014,O`Rourke et.al 2014). Moreover,
individuals with higher degrees of intrinsic motivation from a growth mindset exhibit a
significant higher ability for learning and consequently for embracing new learning
opportunities (Yeager et al 2012). In other words, individuals with a growth mindset are
more open to change because it enhances their intrinsic motivation from learning which
fuels their individual utility (Ng 2018). In a situation with considerable change stemming
from execution of dynamic capabilities, an organization with a common cognitive frame
based on a growth mindset will be able to capture more value from the change \emph{et
  ceteris paribus}.

Consequently, by instilling a growth mindset as a change in the cognitive frame of
Microsoft, Nadella enhanced the change ability of the firm by strengthening the intrinsic
motivation employees have to buy into the change, and meanwhile securing a certain
interest alignment. So Nadella did not only change the resources and capabilites of the
company due to a strategic change towards the cloud, he did also instill a growth mindset
to utilize the motivational and non-behavioral object of action discussed above. In other
words, dynamic capabilites can work through two distinctly different paths. The first path
is the familiar process of changing underlying resources and operational capabilities
(\emph{resource orchestration path}). The second is the focal path of this paper where
dynamic capabilities changes the cognitive frames of the organization \emph{cognitive
  programmer path}. This leads us to the following initial hypotheses:

H1: Dynamic capabilities influences the quality? of operational capabilities

H2: Dynamic capabilities influences the WHAT of cognitive frames

The link between operational capabilities enacted through dynamic capabilites, and
competetive advantage is well established in the literature (refs). The main notion is
that resources and capabilites evolved through dynamic capabilites are inimatable and
rare, and hence conducive to the conditions underpinning competetive advantage (refs
barney). Thus our third hypothesis is well known as the first one:

H3: Operational capabilites acts as a mediator between dynamic capabiliites and
competetive advantage.

Similarly, instilling a growth mindset is also leading to competetive advantge BECAUSE???
\todo{Need to discuss this}

And one more hpythoses on how COG influences resources

To make sure that these interests are aligned

DC executed requires changes. These changes will require a simultanous change in cognitive
frames of the firm for N reasons.

1. Sensing and seizing activities typicaly of dynamic capabilities generate possible
agency problems *blyler coff
2. Information from the process is interpreted and differently at different levels in the
corporate hierarchy (gavetti) hence aligning interpretation across the organization matter

One way of overcoming this agency problems and varying interpratations of the environment,
is to develop a common cognitive frame within which all information is interpreted and
acted upon. Such common traits purpose and organizational identity, 
organizatioanl values and culture \citep{Verona2011}.

To see how dynamic capabilities acts through cognitive objects of action (i.e. as a
cognitive programmer) we consider two distinctly different functions of dynamic
capabilities related to their microfoundations \citep{Teece2007}. The first function is
related to the discovering and capturing opportunities (e.g. the ability of the
organization to sense and seize opportunities (ibid)) which is very much related to the
concept of exploration \citep{March1991}. Organizations with high levels of dynamic
capabilities geared towards sensing and seizing opportunities are doing so through
\emph{resource orchestration}, e.g. by creating, extending and modifying the resource
base, as discussed extensively in the extant literature
\citep{Helfat2007,DiStefano2014,Protogerou2012,Schilke2018}. However, sensing and seizing
opportunities in the market relates to discovering market disequilibrium
\citep{Kirzner1997} as well as upsetting existing equilibria \citep{Schumpeter1934}. The
process of sensing and seizing 'involves understanding latent demand, the structural
evolution of industries and markets, and likely supplier and competitor responses'
\citep[p. 1322]{Teece2007} as well as 'a good deal of customer, competitor, and supplier
information and intelligence (with a) a significant tacit component' (ibid p
1330). Consequently, in this process of exploration individuals in the organization may
accumulate private and tacit information that, if unchecked, can become a source of
knowledge asymmetries and incentive problems (alchian and demsetz 1972), and leave the
firm vulnerable to agency problems. Accumulation of private information in the
organization without incentive alignment of some sort can lead rent-appropriation from
information asymmetries \citep{Blyler2003}. One way to  if not sufficiently aligned with the
cognitive frames of the organization. This particular

%Mindset induction effects on cognitive control: a neurobehavioral investigation.
%Schroder HS, Moran TP, Donnellan MB, Moser JS
%Biol Psychol. 2014 Dec; 103():27-37.

%12. Yeager D.S., Dweck C.S. Mindsets that promote resilience: When students believe that personal characteristics can be developed. Educ. Psych%ol. 2012;47:302–314

Proactiveness is an opportunity-seeking, forward-looking perspective characterized by the introduction of new products and ser- vices ahead of the competition and acting in anticipation of future demand.

Proacteivenss can create psychological ownership (siegel 
We build on these theoretical considerations by applying the concept of psychological ownership (Pierce et al., 2003), defined as ‘the state in which individuals feel as though the target of ownership or a piece of that target is “theirs” ’ (Pierce et al., 2003, p. 86)

Missing link between growth mindset and internal motivation
https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5836039/

Growth mindset and individual proactivity
https://www.emerald.com/insight/content/doi/10.1108/CDI-11-2016-0194/full/html

To be clear, the importance of motivation and, in particular, its strategic relevance for
organi- zational performance and competitive advan- tage has surfaced in some recent
contributions in the strategic management field. Managerial rent theorists (Castanias &
Helfat, 1991, 2001), for example, submit that the availability of skills that have rent
generation potential is a neces- sary but not sufficient condition for rent gener-
ation. In so doing, they also point to the impor- tance of motivation for the realization
of any such potential. Similarly, Coff (1997) discusses a number of dilemmas firms face in
generating an advantage from human assets, and he points to the importance of governance
mechanisms to overcome them. Finally, Makadok (2003) calls for future research into the
genesis of competitive advantage in a way that combines competence- based issues with
questions of governance or motivation.

of the
company’s existence and what defi nes or limits its role(s) within its social and eco-
nomic environment (organizational identity)? What does it stand for (organizational val-
ues)? How are things done over here (organizational culture)?

include norms (leonard Barton 1992), avoidance of cognitive
inertia (Trpisas and Gavetti 2000)

organizational inertia
Second, 
The existence of layer upon layer of standard procedures, established capabilities, complementary assets, and/or admin- istrative routines can exacerbate decision-making biases against innovation. Incumbent enterprises,

Individuals are the unit of analysis embedding these abilities and if left up to
these individuals without a higher organizing framework, the firm will be very vulnerable
because of knowledge asymmetries and incentive problems (alchian and demsetz 1972) and
lead to sub optimal rent-appropriation on the part of the individual employee
\citep{Blyler2003}. A similar argument applies to the function of \emph{seizing} where the
organization is set to mobilize resources for the purpose of gaining from these
opportunities. In order for value to be captured from opportunities, a string of
conditions have to be met \citep{Teece2017}, but the evolution of a 'strong and
change-oriented organizational culture' \citep[p. 331]{Teece2014} is of pivotal
importance. Alas the sensing process requires new knowledge from a wide range of sources
in the organization a lot of this tacit knowledge can lead to rent-appropriation
from information asymmetries if not sufficiently aligned with the
cognitive frames of the organization \citep{Blyler2003}. Furthermore, as cognition plays
an important role in the sensing and seizing of opportunities \citep{Helfat2015} and the
interpretation and understanding varies with individuals position within the
organization, then diverging interpretations stemming from misaligned cognitive frames may
lead to

They maintain that social capital and notably individuals’ valuable internal and external social ties allow for information sharing, innovation and novel ways of thinking which in turn helps managers understand resource acquisition, integration and release

Although they relate the existence of social capital
as an antecedents to dynamic capabilities, the same argument can be applied to their
function; to be able to capture rents for the organization and not the individual a
certain level of social capital is needed and

“firms would be unable to acquire, recombine, and release resources”
resources” (2003: 680) without the social capital of individuals. Hence need to build
social capital. Growth mindset - cognitive frames important to build social captail within
the organization?

@@

the emphasis is on the intermediary role of "asset
orchestration" \citep{Teece2012,Sirmon2011} as related to the "the selection, con-
figuration, alignment, and modification of tangible and intangible assets"
\citep[p.842]{Helfat2015}. This process is very much similar to the exploration and
exploitation of opportunities and we consider reconfiguration to come into the focus when
these processes are seen over time and in changing contexts. Hence we are limiting this
study to looking at the sensing and seizing class of dynamic capabilities.

@@
